---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
>Show any code that is needed to

 1. Load the data (i.e. read.csv())
 
 2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r echo=TRUE}
data <- read.csv("activity.csv")
```

A portion of the original dataset is as follows:
```{r echo=FALSE}
print(data[1:20,])
```

### What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in
the dataset.

1. Make a histogram of the total number of steps taken each day

2. Calculate and report the **mean** and **median** total number of steps taken per day


```{r echo=TRUE}
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
```


1. Mean
```{r echo=TRUE}
      rmean <- mean(steps_by_day$steps,na.rm = TRUE)
```
```{r echo=FALSE}
      print(rmean)
```
2. Median
```{r echo=TRUE}
      rmedian <- median(steps_by_day$steps)
```
```{r echo=FALSE}
      print(rmedian)
```



## What is the average daily activity pattern?

* Calculate average steps for each interval for all days. 
* Plot the Average Number Steps per Day by Interval. 
* Find interval with most average steps. 
```{r}
steps_by_interval <- aggregate(steps ~ interval, data, mean)

plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```

Maximum number of steps interval :  **`r as.character(max_interval)`**  to  **`r as.character(max_interval+5)`** minutes.


## Impute missing values. Compare imputed to non-imputed data.
Missing data needed to be imputed. Only a simple imputation approach was required for this assignment. 
Missing values were imputed by inserting the average for each interval. 

```{r}
incomplete <- sum(!complete.cases(data))
step_i <- data.frame(data$steps)
step_i[is.na(step_i),] <- ceiling(tapply(X=data$steps,INDEX=data$interval,FUN=mean,na.rm=TRUE))
imp_data <- cbind(step_i, data[,2:3])
colnames(imp_data) <- c("steps", "date", "interval")
```


Recount total steps by day and create Histogram. 
```{r}
steps_by_day_i <- aggregate(steps ~ date, imp_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
```

Calculate new mean and median for imputed data. 
```{r echo=TRUE}
rmean_i <- mean(steps_by_day_i$steps)
rmedian_i <- median(steps_by_day_i$steps)
```

Imputed Mean
```{r echo=FALSE}
     
      print(rmean_i)
```

Imputed Median
```{r echo=FALSE}
 
      print(rmedian_i)
```
Calculate difference between imputed and non-imputed data.
```{r}
mean_diff <- rmean_i - rmean
mean_diff
med_diff <- rmedian_i - rmedian
med_diff

```

Calculate total difference.
```{r}
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
total_diff
```


## Are there differences in activity patterns between weekdays and weekends?


1.  A new column indicating whether the date is a weekday or a weekend is added to the new dataset created in the previous section.

```{r echo=TRUE}
  DayType <- data.frame(sapply(X=imp_data$date, FUN=function(day) {
    if (weekdays(as.Date(day)) %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) {
      day <- "weekday"
    }
    else {
      day <- "weekend"
    } 
  }))
  imp_data2 <- cbind(imp_data, DayType)
  
  colnames(imp_data2) <- c("Steps", "Date", "Interval", "DayType")
```

2. The data is then separated into weekday or weekend and the mean (average) number of steps taken for each 5-minute interval, itself averaged across all weekday days or weekend days is calculated.

```{r echo=TRUE, tidy=FALSE}
  dayTypeIntervalSteps <- aggregate(
      data=imp_data2,
      Steps ~ DayType + Interval,
      FUN=mean
  )
```  

A portion of the dataset is as follows:
```{r echo=FALSE}
  print(dayTypeIntervalSteps[1:20,])
```


. Finally, a panel plot of both weekend and weekday graphs is generated.

```{r fig.width=15, echo=TRUE, tidy=FALSE}
  library("lattice")
  
  xyplot(
      type="l",
      data=dayTypeIntervalSteps,
      Steps ~ Interval | DayType,
      xlab="Interval",
      ylab="Number of steps",
      layout=c(1,2)
  )
```